Algebraic Groups\vspace{-1ex}

### 2019-10-21 Algebraic Geometry/Algebraic-Geometry.tex
update

\subsection{Representable or not?}
\begin{itemize}
	\item $\Ga$ as a functor from $\SpcFun^{op}\to\Set$ is represented by $\A1$.
	\item $\Gm$ as a functor from $\SpcFun^{op}\to\Set$ is represented by $\A1\setminus\{0\}$.
	\item $\GL_n:\SpcFun^{op}\to\Set$
	\item $|\cdot|:\SpcFun\to\Set$ underlying set 
	\item $\varnothing:\SpcFun^{op}\to\Set$ sending $X\mapsto\varnothing$
	\item $\{\varnothing\}:\SpcFun^{op}\to\Set$ sending $X\mapsto\{\varnothing\}$.
\end{itemize}

>>>>

\subsection{Representable or not?}
\begin{itemize}
	\item $\Ga$ as a functor from $\SpcFun^{op}\to\Set$ is represented by $\A1$.
	\item $\Gm$ as a functor from $\SpcFun^{op}\to\Set$ is represented by $\A1\setminus\{0\}$.
	\item $\GL_n:\SpcFun^{op}\to\Set$ which is represented by $\Spec k[x_{ij}]_{\det}$
	\item $|\cdot|:\SpcFun\to\Set$ underlying set represented by $\ast$
	\item $\varnothing:\SpcFun^{op}\to\Set$ sending $X\mapsto\varnothing$. Not representable!
	\item $\{\varnothing\}:\SpcFun^{op}\to\Set$ sending $X\mapsto\{\varnothing\}$. Representable by $\ast$.
	\item $\mu_n:\SpcFun^{op}\to\Set$ sending $X\mapsto\{f\in\O_X(X)|f^n=1\}$ represented by $\Spec k[t](t^n-1)$
\end{itemize}

\section{October 18th, 2019}
This is something that Max just mentioned: what is $\O_X(\varnothing)?$ There is a single map here!

\subsection{Old Questions}
Is $U=\A2\setminus (0,0)$ affine? Well first notice 
\[\O_{U}(U)\subseteq \bigcap_{f\in k[x,y]\text{ irred}}k[x,y][1/f]=k[x,y]\]
\textbf{Why is the last equality true?! We're working over a UFD.} But then of course 
$U$ can't be affine since $U\ne \A2=\Spec k[U].$ Another way to think of this is that every isomorphism of 
coordinate rings yields an isomorphism of spaces!

\subsection{Back (again) to affine varieties}
\begin{thm}
	\begin{itemize}
		\item Let $\calA\subseteq\Alg_k$ be the subcategory of finitely-generated, reduced $k$-algebras. Then $\Spec$ defines an equivalence of categories
		\[\Spec:\calA^{op}\to \mathbf{Aff}\]
		\item The functor
		\[X\mapsto\Spec \O_X(X):\SpcFun^\ast\to\mathbf{Aff}\]
		is left adjoint to the canonical inclusion $\mathbf{Aff}\subseteq\SpcFun^\ast$, where $\SpcFun^\ast$
		is the subcategory of $\SpcFun$ such that the global sections are finitely generated.

		Thus $X\to\Spec \O_X(X)$ is universal for maps to affine varieties.
	\end{itemize}	
\end{thm}
\begin{prob}
	The last thing is saying that every morphism $X\to \Spec A$ factors (uniquely) through $X\to\Spec \O_X(X)$. Show this!
\end{prob}

\section{October 21st, 2019}
Today we are moving on from Yoneda to speak a bit about the topological properties of algebraic varieties.

Later we will do a lot more examples with the functor of points perspective, but for now the idea to keep around 
is that this enables us to study something \textit{in relation to something else}.

\subsection{Questions}
We thought about $\mu_n:\SpcFun^{op}\to\Set$. Notice that (at least when $\ch k\nmid n$), we get that 
$k[t]/(t^n-1)$, the representing algebra, is $\prod_{\zeta\in\mu_n(k)}k$. Probably the best way to understand 
this is that it is $n$ points, but moreso that it is embedded in a natural way in $\Gm$ via the SES:
\[\mu_n\hookrightarrow\Ga\to\Gm\]

\subsection{Topological properties of varieties}
\begin{defn}
	A topological space $X$ is \textbf{quasi-compact} if every open cover has a finite subcover.
\end{defn}
\begin{rmk}
	In France, compactness requires a space be Hausdorff. Thus in the development of AG we used the French 
	definition and it stuck!
\end{rmk}
\begin{lem}
	An affine variety is quasi-compact.
\end{lem}
\begin{prf}
	The idea was that if $X=\cup U_i$, then since $D(f)$ generate the topology for $X$, we can refine
	the cover so that $U_i=D(f_i)$. By the Nullstellensatz, $X=\cup D(f_i)$, we know that $(f_i)\lhd \O_X(X)$
	is the whole ring!

	But then $1=\sum_{j=1}^n a_jf_{i_j}$, so we can restrict to a finite cover!
\end{prf}
But note that we didn't need that our ring was Noetherian! So the exact same proof shows that $\Spec \bbC[x_1,x_2,\dots]$ is 
quasi-compact! But $\Spec \bbC[x_i]/(x_1,\dots,x_n,\dots)$ is not quasi-compact!!!

\begin{rmk}
	A very cool (``AMAZING'' according to Max) fact is the following: if $U\subset X$ is any open in an affine space, then $U$ is quasi-compact.
	Compare a similar idea in Hausdorff spaces, where this is actually false.
\end{rmk}

Let $X=\Spec A$ where $A=k[x_1,\dots,x_n]/(f_1,\dots,f_m)$, which is Noetherian by Hilbert basis and correspondence theorem.
Now we have an anti-isomorphism of lattices between the radical ideals in $A$ and the closed sets in $X$.
Thus 
\begin{defn}
	A topological space $X$ is Noetherian if any \textbf{descending} chain of closed subspaces stabilizes.
\end{defn}
\begin{rmk}
	Assuming the axiom of choice, this is equivalent to the statement ``A non-empty set of closed subsets in $X$ has 
	a minimal element.''
\end{rmk}
\begin{rmk}
	Note that $\Spec A$ is Noetherian for any (finitely-generated) $k$-algebra.
\end{rmk}
So let $X$ be a Noetherian topological space.
\begin{lem}
	$Z\subseteq X$ is closed implies that $Z$ is Noetherian and quasi-compact.
\end{lem}
\begin{rmk}
	The Noetherian bit is pretty clear! The quasi-compact part needs a bit of arguing. Use the definition that $X$ is quasi-compact 
	if and only if $\cap Z_i=\varnothing\twoheadrightarrow \cap_{j=1}^n Z_{i_j}=\varnothing$ for some $i_j$.

	Then using choice and the contrapositive it falls out!
\end{rmk}
\begin{defn}
	A topological space $X$ is \textbf{irreducible} if $X\ne\varnothing$ and if $X=X_1\cup X_2$ where 
	the $X_i$ are both closed imples that either $X=X_1$ or $X=X_2$.
\end{defn}
\begin{rmk}
	Equivalently, if $U_1$ and $U_2$ are open and nonempty then $U_1\cap U_2\ne\varnothing$.
	We also have that any non-empty open set is dense.
\end{rmk}

### 2019-10-21 Algebraic Groups/Algebraic-Groups.tex
Thanks Sean!
update

% Document-Specific Macros

@@

Consider $\bbZ/p$ in $\ch p$. We get an action $\bbZ/p$ on $k^2$ via 
\[1\cdot(x,y)=(x+y,y).\]
But notice that $k\stackrel{\hookrightarrow}{y=0}k^2$ is a subrepresentation, but has no complement! Thus this group is not linearly reductive!

@@

>>>>

% Document-Specific Macros
\DeclareMathOperator{\SL}{SL}

@@

Consider $\bbZ/p$ in $\ch p$. We get an action $\bbZ/p$ on $k^2$ via 
\[1\cdot(x,y)=(x+y,y).\]
But notice that $k\stackrel{y=0}{\hookrightarrow}k^2$ is a subrepresentation, but has no complement! Thus this group is not linearly reductive!

@@

\section{October 18th, 2019}
Last time we say that any linear algebraic group $G$ embeds into $\GL_n$. The argument was basically that you look at the global functions 
$\Gamma(G)$ and doing cool stuff. Right at the end Taffy and Tuomas figured out that we just needed to use the other ``side'' of the counit diagram.

\subsection{An example}
Consider 
\[\operatorname{PGL}_2=(\operatorname{Proj} k[a,b,c,d])_{ad-bc]}=\Spec(k[a,b,c,d]_{\det})_0\]
Then consider the representation spanned by 
\[\left\langle\frac{a^2}{\det},\frac{ab}{\det},\dots,\frac{d^2}{\det}\right\rangle\]
which has dimension 10 in $\Gamma(\operatorname{PGL}_2)$.

Thus we have a representation $\operatorname{PGL}_2\to\GL_{10}$. We can compute the matrix representing a matrix (whose determinant
can be assumed to be 1 since we are modding out by scalars).
\begin{prob}
	Do this! In Sage or something.
\end{prob}
\subsection{Special Linear Groups}
Lets discuss $\SL_2$. 
\begin{thm}
	If $\ch k=0$, then 
	\begin{itemize}
		\item $\SL_2$ is linearly reductive.
		\item Every irreducible representation of $\SL_2$ is isomorphic to $\Sym^d k^2$ for some $d$.
	\end{itemize}
	where $k^2$ is the standard representation of $\SL_2$.
\end{thm}
\begin{prf}
	\textit{Sketch:} Recall that in the proof of Maschke one takes a surjection $V\twoheadrightarrow W$ 
	of $G$ representations and we want to show it has a section. We pick a section $s$ in terms of vector spaces and then ``average''
	it:
	\[\tilde s:W\to V\qquad w\mapsto\frac{1}{|G|}\sum_{g\in G}g\cdot s(g^{-1}w)\]
	which is our section.

	Then using the Harr measure on the group, we get from the inclusion of (compact) $\operatorname{SU}_2$ in $\SL_2$ with quotient $\bbC$
	and we can construct the section via 
	\[w\mapsto \int_Gg\cdot s(g^{-1}w)\,\mathrm{d}g\]
	and we get $T_e\SL_2=(T_e\operatorname{SU}_2\otimes_\bbR\bbC)$ and then there is a bit more Lie theory needed to show this makes full sense over $\bbC$.
\end{prf}

Now consider $\Sym^d(k^2)^\vee$, the degree $d$ polynomials on two variables. Then we get an action of $\SL_2$ via 
\[g\cdot f(x,y)=f(g^{-1}\binom{x}{y})-f(dx-by,-cx+ay)\]
Then, as many arguments in linear algebraic groups, we can reduce to a so-called \textit{maximal torus} 
of matrices $(\begin{smallmatrix}a & 0\\0 & a^{-1}\end{smallmatrix})\cong\Gm$. Then we can use techniques on the lie algebra 
$T_e\SL_2=\sl_2$.

Whoa coool. The short exact sequence 
\[1\to \SL_2\to \GL_n\xrightarrow{\det}\Gm\to 1\]
gives us that any representation of $\SL_2$ lifts via $-\otimes \det^i$ to a representation of $\GL_2$.
Sean asked whether this actually gets all the representations or just the polynomial ones. I feel like I should know the 
answer to this. Jarod seems to think this is all of them. I think \href{https://math.stackexchange.com/questions/694001/non-polynomial-representations-of-gl-n}{this stack post}
says something about that.

\textit{I got caught up in thinking and googling and missed an example.}

Let $\ch k=p$ and let $\alpha\in k^\times$. Then define $\alpha\cdot f=\alpha^pf$. This gives us a map 
\[\Sym^d k^n\to \Sym^{dp}k^n\]
that is additive taking $p^{th}$ powers.

So then $\Sym^N k^n$ is \textbf{not simple} if $p|N$.

\section{October 21st, 2019}
Recall that we had that any linear algebraic group over a field $k$ injects into $\GL_n$ as a closed subgroup for some $n$.

An open question is as follows: if $G\to\Spec k[\varepsilon]/\varepsilon^2$ is flat affine group scheme of finite type
Then is $G\subseteq\GL_{n,k[\varepsilon]}$ for some $n?$ This question was asked (as far as Jarod knows) by Brian Conrad on Stack Overflow 
and is still open.

Our goal is to answer the following: if $\phi:H\to H$ then $K=\ker\varphi=H\times_G\Spec k\subseteq H$ is a closed subgroup. What about its image $H/K$?

\subsection{Torsors}
Today we are going to be talking about $G$-torsors.
\begin{defn}
	If $G$ is a group, a \textbf{torsor under $G$} is a set $P$ with a free and transitive group action.
\end{defn}
\begin{rmk}
	So then by fixing a point $p\in P$, we get $G\xrightarrow{\sim}P$ by sending $g\mapsto pg$. In this way, it is like thinking about 
	a group without the identity.
\end{rmk}
An example is by taking a Galois extension $K(\alpha)/K$ with minimal polynomial $f$ of $\alpha$. Then $G=\operatorname{Gal}(K(\alpha)/K)$ acts on 
$\{x|f(x)=0\}$, which is a $G$-torsor.
\begin{defn}
	A \textbf{$G$-torsor over a set $S$} is a set $P$ with a free right $G$-action such that $P\to S$ is $G$-invariant and $S\cong P/G$.
\end{defn}
\begin{rmk}
	Notice that a torsor under $G$ is a specialization of this definition by requiring that $S=\{\ast\}$, the singleton set.
\end{rmk}
\begin{ex}
	Let $H\subseteq G$. THen $H$ acting on $H\to H\setminus G$ (left cosets $gH$ of $H$) is an $H$-torsor.
\end{ex}
\subsection{Flatness}
\begin{defn}
	A map of rings $A\to B$ is \textbf{flat} if $-\otimes_A B$ is exact.
\end{defn}
\begin{rmk}
	Equivalently: for all $p\in \Spec A$, $A_p\to B_p$ is flat. Also: for all $q\in\Spec B$. $A_{\phi^{-1}(q)}\to B_q$ is flat.
\end{rmk}
\begin{defn}
	$A\to B$ is \textbf{faithfully flat} if and only if $-\otimes_AB$ is \textbf{faithfully exact} (exactness and its converse).
\end{defn}
\begin{rmk}
	Other equivalence to faithful flatness: $A\to B$ is flat and $\Spec B\twoheadrightarrow\Spec A$; or $A\to B$ 
	is flat and for any $A$-module $M$, $M=0\leftrightarrow m\otimes_AB=0$.
\end{rmk}
\begin{rmk}
	If $\Spec B\twoheadrightarrow\Spec A$ is faithfully flat and finite presented, then $\Spec A$ has the quotient topology.
\end{rmk}

\begin{prop}
	Let $S=\Spec A$ be Noetherian. Let $G\to S$ be an affine group scheme of flat and finite type over $S$. Let $P\to S$ 
	be a scheme over $S$ with a right $G$-action $P\times_SG\xrightarrow{\sigma}P$. Then the following are equivalent:
	\begin{enumerate}
		\item $P\to S$ is affine, (faithfully?)\footnote{We tried to prove this in class and it seemed not to be true if we don't say this}
		flat, finite type and $(\sigma,\pi_P):(p,g)\mapsto (pg,p)$ is an isomorphism.
		\item There exists a faithfully flat $S'$ such that 
		\begin{center}
			\begin{tikzcd}
				P_{S'}\ar[r]\ar[d] & P\ar[d]\\
				S'\ar[r,"\sigma"] &S
			\end{tikzcd}
		\end{center}
		And $P_{S'}\cong G_{S'}$ as $G_{S'}$-modules.
	\end{enumerate}
\end{prop}
\begin{rmk}
	Note that the above says exactly that $P$ is a $G$-torsor. Another name that has been mentioned and Jarod seems to 
	like is \textbf{principal $G$ bundle.}
\end{rmk}

\subsection{Descent}
Along the way of proving the above propositon, we use the idea of descent.
\begin{lem}
	Consider $X'=\Spec B'\stackrel{\text{f.flat, f.type}}{\twoheadrightarrow}X=\Spec B\to Y=\Spec A$
	where $X',X,$ and $Y$ are Noetherian. Then 
	\begin{enumerate}
		\item $X'\to Y$ is flat implies that $X\to Y$ is flat.
		\item $X'\to Y$ is faithfully flat implies that $X\to Y$ is faithfully flat.
		\item $X'\to Y$ is finite type implies that $X\to Y$ is.
	\end{enumerate}
\end{lem}
\begin{rmk}
	The idea for the first two is just looking at the functors using that $B\to B'$ is faithfully flat. For the third, if $B=\cup_\lambda B_\lambda$,
	then $A\to B_\lambda$ is finitely generated. Then tensoring over $B$ with $B'$ gets us $B'=\cup_\lambda B_\lambda\otimes_B B'$.

	But since $A$ is finitely generated over $B_\lambda$< eventually $B_\lambda\otimes_BB'=B'$. Then 
	consider
	\[0\to B_\lambda\to B\to B/B_\lambda.\]
	After tensoring with (faithfully flat!) $B'$ over $B$, since for some $\lambda$
	\[0\to B_\lambda\otimes_B B'\xrightarrow{\sim}B'\to B/B_\lambda\otimes_BB'\to 0\]
	is exact, forcing the rightmost term to be zero. But by faithfulness this implies $B/B_\lambda=0$ and we are done.
\end{rmk}
\begin{prop}
	Consider
	\begin{center}
		\begin{tikzcd}
			X'\ar[r]\ar[d] & X\ar[d]\\
			S'=\Spec A'\ar[r,"\text{f.type, f.flat}",two heads,swap] & S=\Spec A
		\end{tikzcd}
	\end{center}
	which is a Cartesian square. Then 
	\begin{enumerate}
		\item $X'\to S'$ is an isomorphism iff $X\to S$ is.
		\item $X'\to S'$ is affine iff $X\to S$ is.
	\end{enumerate}
\end{prop}
